---
title: "Introduction to ggbrace"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ggbrace}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The purpose of ggbrace is to produce curly braces for ggplot2. It provides three methods to create curly braces in ggplot2 via two functions: geom_brace and stat_brace.

- geom_brace in default mode plots a brace based on x and y values manually provided.
- geom_brace can plot braces based on the confines of data when the mapping or `inherit.aes` arguments are given.
- stat_brace plots braces partially in the confines of data so that the brace is set apart from it.

For this introduction we will use the mtcars data set to create a dotplot. Then we look at how each of the three different modes draws braces to that plot.

```{r setup}
library(ggbrace)
library(ggplot2)
plt <- ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width, color=Species, label=Species)) + 
  geom_point() +
  theme_classic() +
  theme(legend.position="none")
```

```{r}
plt + stat_brace()
```


```{r}
plt + 
  stat_brace()+  
  stat_bracetext()
```

# Labels & Modifications

## Labels

We can modify the text in the same way we would modify other text in ggplot. We can also switch between different text geoms (e.g. `geom=label`).

```{r}
plt + 
  stat_brace()+  
  stat_bracetext(size=6, angle=15, fontface="bold")

plt + 
  stat_brace()+  
  stat_bracetext(geom="label")
```

## Rotation

We can rotate the braces by 90, 180 or 270 degrees via the `rotate` arguement. The labels are not automatically rotated. For that we have to define the `angle` argument separately

```{r}
plt + 
  stat_brace(rotate=90)+  
  stat_bracetext(rotate=90)
```


```{r}
plt + stat_brace(rotate = 90)
```

## Location

By default, the location of the brace is beside the data points. We can define how far away, where and how big the braces are. We can also define the `bending`, i.e. the curvature. This last parameter can also be set in `geom_brace` (not shown here).

```{r}
plt + stat_brace(outside = FALSE) # the braces are put at a defined distance to the last data point of their group
```

By default, braces have a position and shape that is calculated based on their data points.The position can be changed with the parameters `distance` (to the data points) and `outerstart` (in the coordinate system). The width of the braces can be set with the `width` argument (absolute coordinate system units), while the bending of the brace can be set with `bending` (number from 0 to 1).

- all braces are put at the same distance to their data points

```{r}
plt + stat_brace(distance = 2)
```

- all braces are put at the same position

```{r}
plt + stat_brace(outerstart = 5)
```

- all braces get the same width

```{r}
plt + stat_brace(outerstart = 5, width = 1)
```

- all braces get the same curvature

```{r}
plt + stat_brace(outerstart = 5, width = 1, bending = 0.1)
```

# Outside of plotting area

To vizualize the brace outside of the plotting area, we can simply use two ggplot2 functions. 
- `coord_cartesian` needs to be mentioned with x and/or y range of the plotting area and the parameter `clip="off"` to allow plotting of objects outside of that area.
- within the `theme` function, `plot.margin` needs to be set to expand the outside area. This happens with 4 numbers (above, right, below, left). This can best be achieved with the `"npc"` unit, as it reflects the plot units and is therefore maintained independent of the pixels, cm, etc. Other units can be seen with ?unit (under grid).

```{r}
plt + stat_brace(outerstart = 4.5) + 
  coord_cartesian(y=range(iris$Sepal.Width), clip = "off") + #for the range just use the data for the respective axis
  theme(plot.margin = unit(c(0.25, 0.11, 0.11, 0.11), units="npc"))
```

# Discrete values

Unfortunately, as of now, ggbrace isn't behaving well with discrete x/y axes, which is why they will have to be wrapped into the `seq_along` function within the `aes()`.

```{r}
df <- data.frame(x = c("a","b","c","d","e"), y = 1:5)     

ggplot(df, aes(x, y)) +
  geom_point() +
  stat_brace(aes(x=seq_along(x)))
```

This wrapping into the `seq_along` function is also used in the `coord_cartesian` function when trying to plot outside the plotting area.

```{r}
df <- data.frame(x = c("a","b","c","d","e"), y = 1:5)     

ggplot(df, aes(x, y)) +
  geom_point() +
  stat_brace(aes(x=seq_along(x)), rotate=90) +
  coord_cartesian(x=range(seq_along(df$x)), clip = "off") + 
  theme(plot.margin = unit(c(0.5, 7, 0.5, 0.5), units="lines")) #other units would be "cm" etc.
```
